\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-10-08 NAR_DRAFT.tex
Minor text updates.

\begin{abstract}
Identifying somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
Mutations at low allelic frequency, which are carried by only a portion of cells within a tumor, are particularly difficult to detect.
Many algorithms have been developed to identify such mutations, but none models a key aspect of tumor biology.
Namely, every tumor has its own profile of mutation types that it tends to generate.
We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and then uses it as a prior for evaluating potential mutations.
We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
Using simulations, we show that adding the \batcave algorithm to MuTect improves variant identification.
It also improves the calibration of posterior probabilities, enabling a more principled tradeoff between precision and recall.
Lastly, we show that \batcave also performs well on real data.
Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
More broadly, the algorithm can be added to other variant callers, and it can be extend to include additional biological features that affect mutation generation.
\end{abstract}

@@

Cancer develops through the accumulation of somatic mutations and clonal selection of cells with mutations that confer an advantage.
Understanding the evolutionary history of a tumor, including the mutations that drive its growth, the genetic diversity within it, and the accumulation of new mutations, requires accurate variant identification, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
Low frequency mutations challenge current variant calling methods, because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this challenge increases with sequencing depth.

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper \citep{Larson2012}, JointSNVMix2 \citep{Roth2012a}, Seurat \citep{Christoforides2013}, CaVEMan \citep{Jones2016}, and MuClone \citep{Dorri2019}), allele frequency-based analysis (Strelka \citep{Saunders2012}, LoFreq \citep{Wilm2012}, EBCall \citep{Shiraishi2013b}, deepSNV \citep{Gerstung2012}, LoLoPicker \citep{Carrot-Zhang2017}, and MuSE \citep{Fan2016}), and ensemble and deep learning methods (MutationSeq \citep{Ding2012}, BAYSIC \citep{Cantarel2014}, SomaticSeq \citep{Fang2015}, and SNooPer \citep{Spinella2016}).
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, the mutation profiles differs substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our algorithm first estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

@@

At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is local read depth.
Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
\begin{gather}
  \label{eqn:hypothesis}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
\end{gather}
Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
The maximum likelihood estimate of $\nu$ is simply $\hat{f}$, the number of variant reads divided by the local read depth.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given that hypothesis and the prior probability of that hypothesis. 
Assuming that reads are independent, this is
\begin{equation}\label{eqn:1}
  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
\end{equation}
where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
\begin{equation}
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\},
\end{equation}
and $I$ is the indicator function.
Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
\begin{equation}
  \label{eqn:post_pred}
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
\begin{equation}
  \label{eqn:to_allele}
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
\end{equation}

@@

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \textrm{BSgenome} \citep{Pages2019}, \textrm{GenomicAlignments} \citep{Lawrence2013}, \textrm{VariantAnnotation} \citep{Obenchain2014}, and \textrm{SomaticSignatures} \citep{Gehring2015} fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of alleles.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
We ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

\section{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.

>>>>

\begin{abstract}
Identifying somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
Mutations at low allelic frequency, which are carried by only a portion of cells within a tumor, are particularly difficult to detect.
Many algorithms have been developed to identify such mutations, but none models a key aspect of tumor biology.
Namely, every tumor has its own profile of mutation types that it tends to generate.
We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and mutation rate then uses it as a prior for evaluating potential mutations.
We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
Using simulations, we show that adding the \batcave algorithm to MuTect improves variant identification.
It also improves the calibration of posterior probabilities, enabling a more principled tradeoff between precision and recall.
Lastly, we show that \batcave also performs well on real data.
Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
More broadly, the algorithm can be added to other variant callers, and it can be extended to include additional biological features that affect mutation generation.
\end{abstract}

@@

Cancer develops through the accumulation of somatic mutations and clonal selection of cells with mutations that confer an advantage.
Understanding the evolutionary history of a tumor, including the mutations that drive its growth, the genetic diversity within it, and the accumulation of new mutations, requires accurate variant identification, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \cite{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
Low frequency mutations challenge current variant calling methods, because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this challenge increases with sequencing depth.

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka \cite{Saunders2012}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our \batcave algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

@@

At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls $b$ and their associated quality scores $q$, where $\mathrm{d}$ is local read depth.
Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
\begin{gather}
  \label{eqn:hypothesis}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
\end{gather}
Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
The maximum likelihood estimate of $\nu$ is simply $\hat{f}$, the number of variant reads divided by the local read depth.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given that hypothesis and the prior probability of that hypothesis. 
Assuming that reads are independent, this is
\begin{equation}\label{eqn:1}
  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
\end{equation}
where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads, and $p(m,\nu)$ is the prior.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
\begin{equation}
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\},
\end{equation}
and $I$ is the indicator function.
Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
\begin{equation}
  \label{eqn:post_pred}
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
\begin{equation}
  \label{eqn:to_allele}
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,C=c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
\end{equation}

@@

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.\\
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

\section{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
Because our focus is on evaluating the statistical calling model, we computed all test metrics using only those potential variants that passed MuTect's heuristic filters and entered the statistical model.
